ABM simulating short-term mating behaviors

### 2022-06-06 submission/manuscript.Rmd
minor edits based on feedback

# An Agent-Based Model of Gender Differences in Short-Term Mating Behaviors as a Result of Mating Preferences {-}

@@

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which heterosexual and homosexual men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates only when heterosexual men and women who successfully participated in the mating pool (i.e., those with a non-zero number of short-term mating experiences) were considered. Moreover, when men (vs. women) had stronger preferences for short-term mating, gay men had a higher average number of both experiences and mates compared to both lesbian women and heterosexual men. These results suggest that theoretically speaking, even when gender differences in mating preferences exist, those in short-term mating behaviors only occur among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

*Keywords:* agent-based modeling, short-term mating, casual sex, sexual strategy theory, female choice hypothesis

@@

As a starting point, we took gender differences in mating preferences as conceptualized by the sexual strategy theory [@buss_sexual_1993; @buss2019mate], which is repeatedly supported by empirical investigations [e.g., @schmitt_universal_2003; @walter_sex_2020]. Using an agent-based model, we investigated whether men and women's differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates.

@@

Gender differences in mating preferences have been studied in the light of sexual strategy theory [@buss_sexual_1993; @buss2019mate]. It posits that since the minimum obligatory investment that men must devote to their offspring (contribution of sperm through one sexual act) is lower than that of women (gestation, labor, and lactation), men tend to be relatively more interested in short-term mating compared to women. This is because the number of offspring women can produce is limited and cannot be increased by mating with a large number of men, whereas the reverse is the case for men. As a result, men are predicted to not only have a greater interest in short-term mating, but also desire a larger number of short-term mates in a given period and be less selective with respect to accepting a potential mate [@buss_sexual_1993].

The hypotheses derived from sexual strategy theory have received extensive support in empirical studies. For example, men have reported to be currently seeking short-term mates to a larger extent than women [@buss_sexual_1993; @schmitt_are_2001; @schmitt_universal_2003], and a larger proportion of men were in any way seeking short-term mates (vs. not seeking) [@schmitt_universal_2003]. Similar results have been found among both U.S. college students [@buss_sexual_1993; @schmitt_are_2001] as well as in ten world regions in a cross-cultural study [@schmitt_universal_2003]. Similarly, men across the world have also reported to desire more short-term sexual partners compared to women over different future time periods (e.g., a month, a year) [@buss_sexual_1993; @schmitt_universal_2003; @mcburney_preferred_2005]. The gender differences have been found regardless of whether they were estimated using mean [@buss_sexual_1993; @schmitt_are_2001; @schmitt_universal_2003], median [@schmitt_are_2001; @schmitt_universal_2003], or percentage [@schmitt_universal_2003] statistics.

@@

However, it is important to note that although heterosexual men’s total number of short-term mating is equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. Empirical studies have found that the proportion of individuals who are sexual inexperienced is slightly higher among men than women in most age groups above 18 years old [@haydon2014prevalence; @ghaznavi2019trends]. As a postulation, there may be a smaller proportion of men (vs. the proportion of women) who participate in the mating pool and contribute to the total number of short-term mating. For example, suppose that there is a population of 20 individuals with an equal sex ratio. Heterosexual short-term mating has happened \(n\) times in this population, but only 3 men and 8 women have had any short-term mating experiences. Since the total number of short-term mating behaviors is the same between men and women, we would expect that in a population of individuals who had any short-term mating, the average number of short-term mating experiences and mates was higher among men than women (e.g., in the aforementioned example, the average number of short-term experiences is \(\frac{n}{3}\) for men and \(\frac{n}{8}\) for women who are in the mating pool).

As a comparison, in the cases of gay men and lesbian women, men’s preferences are not constrained by women’s, but only by those of other men, who, arguably, have a similar high interest in short-term mating. This would allow for a more direct behavioral expression of men’s mating preferences. The notion that gay men have less restricted preferences is borne out in the proportion of individuals who have engaged in extradyadic sex [@peplau_sexuality_2004]. A study showed that the proportions of heterosexual men and women who had engaged in extradyadic sex were 26% and 21%, respectively, while the statistics among gay men and lesbian women were 82% and 28%, respectively [@peplau_sexuality_2004]. Therefore, we would expect to observe gender differences in short-term mating behaviors between gay men and lesbian women. Moreover, we would also expect to find that gay men, compared to heterosexual men, engaged in more short-term mating behaviors due to less constraints on their mating preferences.

@@

We also modeled forming long-term relationships as a background process. A pair of individuals had a possibility to commit to a long-term relationship upon meeting each other. This process was modeled at a behavioral level and the psychological processes behind it was ignored, although we recognized gender differences may also exist in long-term mating strategy [@buss_sexual_1993; @buss2019mate].

We modeled these processes among both heterosexual individuals and gay men and lesbian women to examine whether constraints set by the opposite sex’s preferences would affect short-term mating behaviors. Individuals’ sexual orientation was conceptualized in terms of behaviors only in our model. Heterosexual men had short-term mating or formed a long-term relationship only with women, while gay men only with other men, and mutatis mutandis for women.

We formulated the following hypotheses in the present study: assuming gender differences in preferences for short-term mating, 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered, 2) a smaller proportion of heterosexual men (vs. heterosexual women) had ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men had higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) gay men would engage in more short-term mating behaviors compared to heterosexual men.

@@

## Comparing across conditions

@@

The present study used agent-based modeling to investigate whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates. We formulated five hypotheses: 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered, 2) a smaller proportion of heterosexual men (vs. heterosexual women) had ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men had higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) gay men would engage in more short-term mating behaviors compared to heterosexual men. The results from 1,000 time steps in a model simulating men and women’s mating behaviors provided strong evidence in favor of our hypotheses. First of all, we found no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered. However, compared to heterosexual women, there was a smaller proportion of heterosexual men participating in the mating pool, but they had higher average numbers of short-term mating experiences and short-term mates. Secondly, as compared to lesbian women, gay men had higher average numbers of short-term mating experiences and short-term mates. Finally, gay men also had higher average numbers of short-term mating experiences and short-term mates as compared to heterosexual men.

@@

Also, as gay men's preferences for short-term mating were not constrained by those of women, we found that gay men engaged in more short-term mating behaviors compared to heterosexual men. This was consistent with the empirical literature [e.g., @peplau_national_1997]. Interestingly, gay men and heterosexual men had the same short-term mating likelihood and the same standards for short-term mates in our model. The only difference was a change in the preferences of potential partners. When men’s partners had a stronger preferences for short-term mating (i.e., having men vs. having women as potential partners), men also appeared to engage in more short-term mating behaviors.

@@

Using agent-based modeling to simulate short-term mating behaviors, we found when men (vs. women) had a stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent when the full population was considered, while men in the mating pool had more such behaviors than women in the mating pool. Gay men engaged in more short-term mating behaviors compared to both lesbian women and heterosexual men.

>>>>

# An Agent-Based Model of Gender and Sexual Orientation Differences in Short-Term Mating Behaviors as a Result of Mating Preferences {-}

@@

Gender differences in short-term mating behaviors are well-documented in human sexuality research. Existing studies usually conflate gender differences in mating preferences with differences in sexual behaviors, which is theoretically problematic. Using an agent-based model, we investigated the circumstances under which men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors. The model showed that when all individuals in a closed heterosexual population were considered,  men and women had the same average number of short-term mating experiences and short-term mates even when men had stronger preferences for short-term mating. Men (vs. women) had a higher average number of both experiences and mates when analyses were limited to only heterosexual men and women who successfully participated in the mating pool (i.e., those with a non-zero number of short-term mating experiences). Moreover, when men (vs. women) had stronger preferences for short-term mating, gay men had a higher average number of both experiences and mates compared to both lesbian women and heterosexual men. These results suggest that even when gender differences in mating preferences exist, those in short-term mating behaviors only occur among particular populations, or when men’s preferences for short-term mating are not constrained by those of women. Suggestions for future research in human mating psychology and behaviors were provided.

@@

*Keywords:* agent-based modeling, short-term mating, casual sex, sexual strategies theory, female choice hypothesis

@@

As a starting point, we took gender differences in mating preferences as conceptualized by the sexual strategies theory [@buss_sexual_1993; @buss2019mate], which is repeatedly supported by empirical investigations [e.g., @schmitt_universal_2003; @walter_sex_2020]. Using an agent-based model, we investigated whether men and women's differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates.

@@

Gender differences in mating preferences have been studied in the light of sexual strategies theory [@buss_sexual_1993; @buss2019mate]. It posits that since the minimum obligatory investment that men must devote to their offspring (contribution of sperm through one sexual act) is lower than that of women (gestation, labor, and lactation), men tend to be relatively more interested in short-term mating compared to women. This is because the number of offspring women can produce is limited and cannot be increased by mating with a large number of men, whereas the reverse is the case for men. As a result, men are predicted to not only have a greater interest in short-term mating, but also desire a larger number of short-term mates in a given period and be less selective with respect to accepting a potential mate [@buss_sexual_1993].

The hypotheses derived from sexual strategies theory have received extensive support in empirical studies. For example, men have reported to be currently seeking short-term mates to a larger extent than women [@buss_sexual_1993; @schmitt_are_2001; @schmitt_universal_2003], and a larger proportion of men were in any way seeking short-term mates (vs. not seeking) [@schmitt_universal_2003]. Similar results have been found among both U.S. college students [@buss_sexual_1993; @schmitt_are_2001] as well as in ten world regions in a cross-cultural study [@schmitt_universal_2003]. Similarly, men across the world have also reported to desire more short-term sexual partners compared to women over different future time periods (e.g., a month, a year) [@buss_sexual_1993; @schmitt_universal_2003; @mcburney_preferred_2005]. The gender differences have been found regardless of whether they were estimated using mean [@buss_sexual_1993; @schmitt_are_2001; @schmitt_universal_2003], median [@schmitt_are_2001; @schmitt_universal_2003], or percentage [@schmitt_universal_2003] statistics.

@@

However, it is important to note that although heterosexual men’s total number of short-term matings would be equal to that of women, this does not necessarily mean that the total number of men and women who have ever had short-term mating must be equal. Empirical studies have found that the proportion of individuals who are sexually inexperienced is slightly higher among men than women in most age groups above 18 years old [@haydon2014prevalence; @ghaznavi2019trends]. Therefore, as a postulation, there may be a smaller proportion of men (vs. the proportion of women) who participate in the mating pool and contribute to the total number of short-term mating. For example, suppose that there is a population of 20 individuals with an equal sex ratio. Heterosexual short-term matings have happened \(n\) times in this population, but only 3 men and 8 women have had any short-term mating experiences. Since the total number of short-term mating behaviors is the same between men and women, we would expect that in a population of individuals who had any short-term mating, the average number of short-term mating experiences and mates was higher among men than women (e.g., in the aforementioned example, the average number of short-term experiences is \(\frac{n}{3}\) for men and \(\frac{n}{8}\) for women who are in the mating pool).

As a comparison, in the cases of gay men and lesbian women, men’s preferences are not constrained by those of women’s, but only by those of other men, who, arguably, have a similar high interest in short-term mating. This would allow for a more direct behavioral expression of men’s mating preferences. The notion that gay men have less restricted preferences is born out in the proportion of individuals who have engaged in extradyadic sex [@peplau_sexuality_2004]. A study showed that the proportions of heterosexual men and women who had engaged in extradyadic sex were 26% and 21%, respectively, while the numbers among gay men and lesbian women were 82% and 28%, respectively [@peplau_sexuality_2004]. Therefore, we would expect to observe gender differences in short-term mating behaviors between gay men and lesbian women. Moreover, we would also expect to find that gay men would engage in more short-term mating behaviors compared to heterosexual men due to less constraints on gay men's mating preferences.

@@

We also modeled forming long-term relationships as a background process. A pair of individuals had a possibility to commit to a long-term relationship upon meeting each other. This process was modeled at a behavioral level and the psychological processes behind it was ignored, although we recognized that gender differences may also exist in long-term mating strategy [@buss_sexual_1993; @buss2019mate].

We modeled these processes among both heterosexual individuals and gay men and lesbian women to examine whether constraints set by the opposite sex’s preferences would affect short-term mating behaviors. Individuals’ sexual orientation was conceptualized as categorical in terms of behaviors only in our model. Heterosexual men had short-term mating or formed a long-term relationship only with women, while gay men only with other men, and mutatis mutandis for women.

We formulated the following hypotheses: assuming gender differences in preferences for short-term mating, 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when considering the full population, 2) a smaller proportion of heterosexual men (vs. heterosexual women) would have ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men would have a higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) compared to heterosexual men.

@@

## Experimental conditions giving rise to gender differences in short-term mating behaviors

@@

The present study used agent-based modeling to investigate whether men and women’s differential preferences for short-term mating would result in gender differences in short-term mating behaviors, and if so, then under what conditions they would result in differences in the number of short-term mating experiences and short-term mates. We formulated five hypotheses: 1) there would be no gender differences in short-term mating behaviors between heterosexual men and women when considering the full population, 2) a smaller proportion of heterosexual men (vs. heterosexual women) would have ever engaged in short-term mating, 3) in a population of heterosexual individuals who had any short-term mating, men would have a higher average number of short-term mating behaviors than women, 4) gay men would engage in more short-term mating behaviors compared to lesbian women, and 5) compared to heterosexual men. The results from 1,000 time steps in a model simulating men and women’s mating behaviors provided strong evidence in favor of our hypotheses. First of all, we found no gender differences in short-term mating behaviors between heterosexual men and women when the full population was considered. However, compared to heterosexual women, there was a smaller proportion of heterosexual men participating in the mating pool, but they had higher average numbers of short-term mating experiences and short-term mates. Secondly, as compared to lesbian women, gay men had higher average numbers of short-term mating experiences and short-term mates. Finally, gay men also had higher average numbers of short-term mating experiences and short-term mates as compared to heterosexual men.

@@

Also, as gay men's preferences for short-term mating were not constrained by those of women, we found that gay men engaged in more short-term mating behaviors compared to heterosexual men. This was consistent with the empirical literature [e.g., @peplau_national_1997]. Interestingly, gay men and heterosexual men had the same short-term mating likelihood and the same standards for short-term mates in our model. The only difference was a change in the preferences of potential partners. When men’s partners had stronger preferences for short-term mating (i.e., having men vs. having women as potential partners), men also appeared to engage in more short-term mating behaviors.

@@

Using agent-based modeling to simulate short-term mating behaviors, we found when men (vs. women) had stronger preferences for short-term mating, heterosexual men and women engaged in short-term mating behaviors to the same extent when the full population was considered, while men in the mating pool had more such behaviors than women in the mating pool. Gay men engaged in more short-term mating behaviors compared to both lesbian women and heterosexual men.
